---
title: "NYPD Crime Project"
author: "Brian Cizek"
output:
  html_document: default
  pdf_document: default
date: "2023-11-27"
---

```{r setup, include=TRUE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(lubridate)
library(plotly)
library(ggplot2)
```


## Get the data

```{r get_nypd_data}
nypd = read.csv("https://data.cityofnewyork.us/api/views/833y-fsy8/rows.csv?accessType=DOWNLOAD")
```


## Clean the data

```{r clean_data}

#Remove the columns that I am not interested in for this project
nypd <- nypd %>% 
  select(-c(Latitude, Longitude, Lon_Lat, X_COORD_CD, Y_COORD_CD))

#Create some new columns that I will use for analysis
nypd <- nypd %>% 
  mutate(OCCUR_DATE = mdy(OCCUR_DATE),
         OCCUR_TIME = hms(OCCUR_TIME),
         OCCUR_HOUR = as.integer(OCCUR_TIME@hour),
         OCCUR_MONTH = as.integer(month(OCCUR_DATE)),
         OCCUR_YEAR = year(OCCUR_DATE),
         STATISTICAL_MURDER_FLAG = as.logical(STATISTICAL_MURDER_FLAG),
         SHOOTINGS = 1,
         MURDER = as.integer(STATISTICAL_MURDER_FLAG))

```

## Handle Missing Data

```{r missing_data}
#Handle missing data

nypd[nypd == ''] <- NA

nypd <- nypd %>%
  replace_na(list(PERP_AGE_GROUP = "UNKNOW", PERP_RACE = "UNKNOWN", PERP_SEX = "UNKNOWN",  VIC_AGE_GROUP = "UNKNOWN", VIC_SEX="UNKNOWN", VIC_RACE="UNKNOWN"))
```


## Borough Analysis

### Barplot for total shootings in each borough

```{r boro_analysis}
#color plot
ggplot(nypd, aes(x=as.factor(BORO), fill=as.factor(BORO))) +
  geom_bar() +
  scale_fill_brewer(palette = "Set1") +
  theme(legend.position="none")
```

### Stacked barplot, split by murders vs. non-murders

```{r boro_stacked_by_murder}
#Barplot by Boro, stacked by murder
ggplot(nypd, aes(BORO, fill = STATISTICAL_MURDER_FLAG)) +
  geom_bar(position = "stack")
```


```{r stacked_percentage}
#Barplot by Boro, stacked by murder percentage
ggplot(nypd, aes(BORO, fill = STATISTICAL_MURDER_FLAG)) +
  geom_bar(position = "fill")
```



## Timeseries Analysis

```{r timeseries}
#Create a new df for time series analysis
nypd_annual <- nypd %>%
  group_by(OCCUR_YEAR, SHOOTINGS)%>%
  summarize(SHOOTINGS = sum(SHOOTINGS),
            STATISTICAL_MURDER_FLAG = sum(STATISTICAL_MURDER_FLAG)) %>%
  select(OCCUR_YEAR, SHOOTINGS, STATISTICAL_MURDER_FLAG) %>%
  ungroup()

nypd_annual_murder <- nypd %>%
  group_by(OCCUR_YEAR, SHOOTINGS)%>%
  summarize(SHOOTINGS = sum(SHOOTINGS),
            MURDER = sum(MURDER)) %>%
  select(OCCUR_YEAR, SHOOTINGS, MURDER) %>%
  ungroup()
  
#Create a new df for boro time series analysis
nypd_annual_boro <- nypd %>%
  group_by(OCCUR_YEAR, SHOOTINGS, BORO)%>%
  summarize(SHOOTINGS = sum(SHOOTINGS),
            STATISTICAL_MURDER_FLAG = sum(STATISTICAL_MURDER_FLAG)) %>%
  select(OCCUR_YEAR, BORO, SHOOTINGS, STATISTICAL_MURDER_FLAG) %>%
  ungroup()


#Timeseries lineplot
ggplot(nypd_annual, aes(x=OCCUR_YEAR)) +
  geom_line(aes(y=SHOOTINGS)) +
  geom_point(aes(y=SHOOTINGS)) +
  geom_line(aes(y=STATISTICAL_MURDER_FLAG), color="red") +
  geom_point(aes(y=STATISTICAL_MURDER_FLAG), color="red") +
  labs(title = "NYPD Shootings (Black) and Murders (Red) by Year",
       x = "Year",
       y = "Shootings/Murders")

#Timeseries by boro
ggplot(nypd_annual_boro, aes(x=OCCUR_YEAR, y=SHOOTINGS, color=BORO)) +
  geom_line() +
  geom_point() +
  labs(title = "NYPD Shootings by Year by Boro",
       x = "Year",
       y = "Shootings")
```


## Time of year and time of day

```{r initial_barplots}

#histogram by hour

ggplot(nypd, aes(x=as.factor(OCCUR_HOUR))) +
  geom_bar(fill="blue") 

#histogram by month

ggplot(nypd, aes(x=as.factor(OCCUR_MONTH))) +
  geom_bar(fill="red") 
```


## 2D Histograms/Contour plots: Shootings by month and time of day

```{r 2d_hist}
#2d histogram by month and time of day
fig1 <- plot_ly(x = nypd$OCCUR_HOUR, y = nypd$OCCUR_MONTH, type = "histogram2dcontour")

fig1 <- fig1 %>% 
        colorbar(title = "Shootings") %>%
        layout(title = 'Time of Day vs. Month', plot_bgcolor = "#e5ecf6", xaxis=list(title = "Time (Local)"), yaxis = list(title = "Month"))

fig1



#putting it all together
s <- subplot(
  plot_ly(x = nypd$OCCUR_HOUR, type = "histogram"),
  plotly_empty(),
  plot_ly(x = nypd$OCCUR_HOUR, y = nypd$OCCUR_MONTH, type = "histogram2dcontour"),
  plot_ly(y = nypd$OCCUR_MONTH, type = "histogram"),
  nrows = 2, heights = c(0.2, 0.8), widths = c(0.8, 0.2), margin = 0,
  shareX = TRUE, shareY = TRUE, titleX = FALSE, titleY = FALSE
)
fig <- layout(s, showlegend = FALSE)
fig
```


## Conclusion and Bias Identification

In this report, I did some analysis on NYPD shooting trends. There appeared to be a steady decline in both shootings and murders until 2020, when there was a sharp increase. Additionally, we examined the time of year and time of day when shootings are most likely to occur. The data show that the summertime and during the evening/nighttime hours are the most dangerous. 

There are a few areas of potential bias in these data. There are likely shootings that go unreported to the NYPD, and hence would not be in this data. There may be some boroughs that report shootings at a higher rate than others. 







